---
title: "Movie Dataset: A K-Means Clustering Analysis"
author: "Shreyes Balaji, Marwan Hegab, Mazin Hussein, Mikael Rotberg, Roberto Rubio, Jordan Woda"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5.5,
  fig.align = "center"
)
```

# 1. Purpose of this report

This report applies K-means clustering to the TMDB 5000 Movie Dataset to
group movies into interpretable clusters based on financial and audience
metrics, then evaluates how well the cluster structure generalizes to a
held-out test set.

# 2. Load packages

```{r packages}
needed_packages <- c(
  "dplyr",
  "tidyr",
  "ggplot2",
  "readr",
  "cluster",
  "factoextra",
  "GGally",
  "knitr",
  "scales",
  "pdist"
)

installed <- rownames(installed.packages())
for (p in needed_packages) {
  if (!(p %in% installed)) install.packages(p)
}

library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(cluster)
library(factoextra)
library(GGally)
library(knitr)
library(scales)
library(pdist)
```

# 3. Read and merge the datasets

```{r read-data}
movies_raw  <- read_csv("data/tmdb_5000_movies.csv", show_col_types = FALSE)
credits_raw <- read_csv("data/tmdb_5000_credits.csv", show_col_types = FALSE)

credits_small <- credits_raw %>%
  select(movie_id, title, cast, crew) %>%
  rename(id = movie_id)

tmdb_raw <- movies_raw %>%
  inner_join(credits_small, by = "id", suffix = c("", "_credits"))

dim(movies_raw)
dim(credits_raw)
dim(tmdb_raw)

head(tmdb_raw[, c("title", "budget", "revenue", "runtime",
                  "popularity", "vote_average", "vote_count")])
str(tmdb_raw)
summary(tmdb_raw[, c("budget", "revenue", "runtime",
                      "popularity", "vote_average", "vote_count")])
```

## 3.1 Variable descriptions

```{r variable-descriptions}
variable_info <- tibble::tibble(
  variable = c(
    "budget", "revenue", "runtime", "popularity",
    "vote_average", "vote_count"
  ),
  description = c(
    "Production budget of the movie in US dollars",
    "Worldwide revenue of the movie in US dollars",
    "Runtime of the movie in minutes",
    "TMDB popularity score (higher = more page views / interactions)",
    "Average user rating (0 to 10 scale)",
    "Number of user votes the movie received"
  )
)
kable(variable_info, caption = "Variable descriptions for the TMDB numeric feature set")
```

## 3.2 Data quality checks

```{r data-quality-checks}
raw_n <- nrow(tmdb_raw)
raw_p <- ncol(tmdb_raw)

na_counts <- colSums(is.na(tmdb_raw[, c("budget", "revenue", "runtime",
                                          "popularity", "vote_average",
                                          "vote_count")]))

duplicate_count <- sum(duplicated(tmdb_raw$id))

quality_checks <- data.frame(
  item = c("Number of rows", "Number of columns", "Duplicated movie IDs"),
  value = c(raw_n, raw_p, duplicate_count)
)
kable(quality_checks, caption = "Basic data quality checks")
kable(
  data.frame(variable = names(na_counts),
             missing_count = as.integer(na_counts)),
  caption = "Missing values for numeric clustering variables"
)
```

## 3.3 Treat zero budget/revenue/runtime as missing

Many TMDB rows store unknown financial values as `0` instead of `NA`,
which would distort distance-based clustering. These are recoded as
missing here.

```{r zero-as-na}
tmdb_raw <- tmdb_raw %>%
  mutate(
    budget = ifelse(budget == 0, NA, budget),
    revenue = ifelse(revenue == 0, NA, revenue),
    runtime = ifelse(runtime == 0, NA, runtime)
  )

zero_na_report <- tmdb_raw %>%
  summarise(
    missing_budget = sum(is.na(budget)),
    missing_revenue = sum(is.na(revenue)),
    missing_runtime = sum(is.na(runtime))
  )
kable(zero_na_report, caption = "Missing counts after recoding zero values as NA")
```

# 4. Raw-data visualizations

## 4.1 Histograms of numeric variables

```{r raw-histograms, fig.height=8}
num_vars <- c("budget", "revenue", "runtime",
              "popularity", "vote_average", "vote_count")

tmdb_long_hist <- tmdb_raw %>%
  select(all_of(num_vars)) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

ggplot(tmdb_long_hist, aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(
    title = "Histograms of numeric variables (raw TMDB data)",
    x = NULL, y = "Count"
  ) +
  theme_minimal()
```

## 4.2 Correlation heatmap for numeric variables

```{r raw-correlation-heatmap}
numeric_raw <- tmdb_raw %>%
  select(all_of(num_vars)) %>%
  drop_na()

corr_mat <- cor(numeric_raw)
corr_df <- as.data.frame(as.table(corr_mat))
names(corr_df) <- c("Var1", "Var2", "Correlation")

ggplot(corr_df, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile() +
  geom_text(aes(label = round(Correlation, 2)), color = "black", size = 3) +
  scale_fill_gradient2(low = "firebrick", mid = "white",
                       high = "navy", midpoint = 0) +
  labs(title = "Correlation heatmap of numeric variables",
       x = NULL, y = NULL, fill = "Corr") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 4.3 Boxplots before cleaning

```{r raw-boxplots, fig.height=8}
ggplot(tmdb_long_hist, aes(x = "", y = value)) +
  geom_boxplot(fill = "tan") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(title = "Boxplots before cleaning", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```

# 5. Clean the data

Cleaning steps used here:

1. Keep only the numeric clustering variables plus `title`
2. Remove rows with missing values (including recoded zeros)
3. Remove duplicate movie IDs
4. Use Tukey's IQR rule on the numeric predictor variables to detect
   and remove outliers

## 5.1 Subset and remove missing values

```{r clean-na-duplicates}
tmdb_step1 <- tmdb_raw %>%
  select(id, title, all_of(num_vars)) %>%
  drop_na()

n_after_na <- nrow(tmdb_step1)

tmdb_step2 <- tmdb_step1 %>% distinct(id, .keep_all = TRUE)
n_after_dups <- nrow(tmdb_step2)

cleaning_counts_1 <- data.frame(
  step = c("Raw data", "After removing NAs", "After removing duplicates"),
  rows = c(nrow(tmdb_raw), n_after_na, n_after_dups)
)
kable(cleaning_counts_1, caption = "Rows remaining after NA and duplicate removal")
```

## 5.2 Detect outliers using Tukey's IQR rule

```{r outlier-detection}
iqr_outlier_flag <- function(x) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower <- q1 - 1.5 * iqr
  upper <- q3 + 1.5 * iqr
  x < lower | x > upper
}

outlier_matrix <- sapply(tmdb_step2[, num_vars], iqr_outlier_flag)
outlier_rows <- apply(outlier_matrix, 1, any)

tmdb_clean <- tmdb_step2[!outlier_rows, ]
n_after_outliers <- nrow(tmdb_clean)

cleaning_counts_2 <- data.frame(
  step = c("After removing duplicates", "Rows flagged as outliers",
           "Final cleaned data"),
  rows = c(nrow(tmdb_step2), sum(outlier_rows), n_after_outliers)
)
kable(cleaning_counts_2, caption = "Rows removed by IQR-based outlier cleaning")
```

## 5.3 Final cleaned dataset checks

```{r cleaned-checks}
dim(tmdb_clean)
summary(tmdb_clean[, num_vars])
```

## 5.4 Boxplots after cleaning

```{r clean-boxplots, fig.height=8}
tmdb_long_box_after <- tmdb_clean %>%
  select(all_of(num_vars)) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

ggplot(tmdb_long_box_after, aes(x = "", y = value)) +
  geom_boxplot(fill = "lightgreen") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(title = "Boxplots after cleaning", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```

# 6. Train-test split

K-means is unsupervised, but the rubric requires a train/test split.
The model is fitted on the training set, and the test set is used to
assign clusters to unseen movies (prediction / evaluation).

```{r split-data}
set.seed(123)

tmdb_features <- tmdb_clean %>% select(all_of(num_vars))

train_index <- sample(seq_len(nrow(tmdb_clean)),
                       size = floor(0.70 * nrow(tmdb_clean)))

train_data <- tmdb_clean[train_index, ]
test_data  <- tmdb_clean[-train_index, ]

split_summary <- data.frame(
  dataset = c("Train", "Test"),
  rows = c(nrow(train_data), nrow(test_data)),
  proportion = c(nrow(train_data) / nrow(tmdb_clean),
                 nrow(test_data)  / nrow(tmdb_clean))
)
kable(split_summary, digits = 3, caption = "Train-test split sizes")
```

## 6.1 Standardize the features

K-means is distance based, so all numeric predictors are standardized
using means and standard deviations from the **training set** only,
then applied to the test set.

```{r scale-features}
train_x <- train_data %>% select(all_of(num_vars))
test_x  <- test_data  %>% select(all_of(num_vars))

train_means <- colMeans(train_x)
train_sds   <- apply(train_x, 2, sd)

train_x_scaled <- scale(train_x, center = train_means, scale = train_sds)
test_x_scaled  <- scale(test_x,  center = train_means, scale = train_sds)

dim(train_x_scaled)
dim(test_x_scaled)
```

## 6.2 Diagonal (pairs) plot to verify assumptions

The pairs / diagonal plot shows the univariate distributions and
pairwise scatterplots of the scaled training features. This is used
to check that the variables are reasonably continuous, to inspect
linear vs. nonlinear pairwise structure, and to look for any
remaining extreme points that could dominate Euclidean distances.

```{r diagonal-plot, fig.height=8, fig.width=9}
set.seed(123)
pairs_sample <- as.data.frame(train_x_scaled) %>%
  slice_sample(n = min(500, nrow(train_x_scaled)))

ggpairs(pairs_sample,
        title = "Diagonal plot of standardized training features") +
  theme_minimal()
```

# 7. Analysis: Choosing the number of clusters K

## 7.1 Elbow method (within-cluster sum of squares)

```{r elbow-plot}
set.seed(123)
fviz_nbclust(train_x_scaled, kmeans, method = "wss", k.max = 10) +
  labs(title = "Elbow method for optimal K (training set)",
       x = "Number of clusters K",
       y = "Total within-cluster sum of squares") +
  theme_minimal()
```

## 7.2 Silhouette method

```{r silhouette-plot}
set.seed(123)
fviz_nbclust(train_x_scaled, kmeans, method = "silhouette", k.max = 10) +
  labs(title = "Silhouette method for optimal K (training set)",
       x = "Number of clusters K",
       y = "Average silhouette width") +
  theme_minimal()
```

## 7.3 Gap statistic (optional confirmation)

```{r gap-plot}
set.seed(123)
gap_stat <- clusGap(train_x_scaled, FUN = kmeans, nstart = 25,
                     K.max = 10, B = 50)
fviz_gap_stat(gap_stat) +
  labs(title = "Gap statistic for optimal K (training set)") +
  theme_minimal()
```

# 8. Fit the final K-means model

The chosen K is set from the elbow and silhouette diagnostics.
Update `final_k` below if a different value is preferred after
inspecting the plots.

```{r kmeans-fit}
final_k <- 4

set.seed(123)
km_final <- kmeans(train_x_scaled, centers = final_k,
                    nstart = 25, iter.max = 100)

km_final$size
km_final$centers
km_final$tot.withinss
km_final$betweenss / km_final$totss
```

## 8.1 Cluster visualization on the first two principal components

```{r cluster-viz}
fviz_cluster(km_final,
             data = train_x_scaled,
             geom = "point",
             ellipse.type = "convex",
             ggtheme = theme_minimal()) +
  labs(title = paste0("K-means cluster plot on training data (K = ",
                       final_k, ")"))
```

## 8.2 Cluster profiles on the original scale

The centers are transformed back from standardized units to the
original scale so the clusters can be interpreted in human-readable
units (dollars, minutes, rating points, etc.).

```{r cluster-profiles}
centers_orig <- sweep(km_final$centers, 2, train_sds, `*`)
centers_orig <- sweep(centers_orig,      2, train_means, `+`)

centers_orig_df <- as.data.frame(centers_orig) %>%
  mutate(cluster = factor(1:final_k),
         size = km_final$size) %>%
  relocate(cluster, size)

kable(centers_orig_df, digits = 2,
      caption = "Cluster centers on the original scale (training set)")
```

## 8.3 Cluster sizes

```{r cluster-sizes}
size_df <- data.frame(
  cluster = factor(1:final_k),
  size = km_final$size
)

ggplot(size_df, aes(x = cluster, y = size, fill = cluster)) +
  geom_col() +
  geom_text(aes(label = size), vjust = -0.3) +
  labs(title = "Training-set cluster sizes",
       x = "Cluster", y = "Number of movies") +
  theme_minimal() +
  theme(legend.position = "none")
```

# 9. Model evaluation and prediction on the test set

## 9.1 Assign each test movie to its nearest training cluster

```{r predict-test}
predict_kmeans <- function(centers, newdata) {
  dist_mat <- as.matrix(
    pdist::pdist(X = as.matrix(newdata), Y = as.matrix(centers))
  )
  apply(dist_mat, 1, which.min)
}

test_clusters  <- predict_kmeans(km_final$centers, test_x_scaled)
train_clusters <- km_final$cluster
```

## 9.2 Test-set cluster sizes and stability

```{r test-cluster-sizes}
train_tbl <- prop.table(table(train_clusters))
test_tbl  <- prop.table(table(test_clusters))

stability_df <- data.frame(
  cluster = factor(1:final_k),
  train_proportion = as.numeric(train_tbl),
  test_proportion  = as.numeric(test_tbl)
)

kable(stability_df, digits = 3,
      caption = "Cluster proportions on the training and test sets")

stability_long <- stability_df %>%
  pivot_longer(cols = c(train_proportion, test_proportion),
               names_to = "set", values_to = "proportion")

ggplot(stability_long, aes(x = cluster, y = proportion, fill = set)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Cluster proportions: train vs test",
       x = "Cluster", y = "Proportion") +
  theme_minimal()
```

## 9.3 Silhouette analysis on the test set

```{r silhouette-test}
test_dist <- dist(test_x_scaled)
sil_test  <- silhouette(test_clusters, test_dist)

summary(sil_test)[["avg.width"]]

fviz_silhouette(sil_test) +
  labs(title = "Silhouette plot on the test set") +
  theme_minimal()
```

## 9.4 Test-set cluster profile on the original scale

```{r test-cluster-profile}
test_with_cluster <- test_data %>%
  mutate(cluster = factor(test_clusters))

test_profile <- test_with_cluster %>%
  group_by(cluster) %>%
  summarise(across(all_of(num_vars), mean), .groups = "drop") %>%
  mutate(size = as.integer(table(test_clusters)))

kable(test_profile, digits = 2,
      caption = "Test-set cluster means on the original scale")
```

## 9.5 Example movies from each cluster

```{r example-movies}
example_movies <- test_with_cluster %>%
  group_by(cluster) %>%
  slice_max(order_by = popularity, n = 5) %>%
  select(cluster, title, budget, revenue, runtime,
         popularity, vote_average, vote_count) %>%
  arrange(cluster)

kable(example_movies, digits = 2,
      caption = "Most popular movies in each test-set cluster")
```

# 10. Summary

```{r auto-summary}
cat("RAW DATA SUMMARY\n")
cat("----------------\n")
cat("Rows in movies data:     ", nrow(movies_raw), "\n")
cat("Rows in credits data:    ", nrow(credits_raw), "\n")
cat("Rows in merged raw data: ", nrow(tmdb_raw),    "\n\n")

cat("CLEANING SUMMARY\n")
cat("----------------\n")
cat("Rows after removing missing values:", n_after_na,       "\n")
cat("Rows after removing duplicates:    ", n_after_dups,      "\n")
cat("Rows flagged as outliers:          ", sum(outlier_rows), "\n")
cat("Rows in final cleaned dataset:     ", n_after_outliers,  "\n\n")

cat("SPLIT SUMMARY\n")
cat("-------------\n")
cat("Train rows:", nrow(train_data), "\n")
cat("Test rows: ", nrow(test_data),  "\n\n")

cat("MODEL SUMMARY\n")
cat("-------------\n")
cat("Final K chosen:                   ", final_k, "\n")
cat("Total within-cluster SS (train):  ",
    round(km_final$tot.withinss, 2), "\n")
cat("Between / total SS ratio (train): ",
    round(km_final$betweenss / km_final$totss, 4), "\n")
cat("Average silhouette on test set:   ",
    round(summary(sil_test)[["avg.width"]], 4), "\n")
```
